Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-01 paper.tex

Zarr is flexible in allowing different compression codecs and 
pre-compression filters to be specified on a per-array basis
(see section XXX).
Two key technologies used by Zarr are the Blosc
% This seems like the best citation for Blosc? It is mentioned here
meta-compressor~\cite{alted2010modern}
and Zstandard compression algorithm~\citep{collet2021rfc}.
Blosc is a high-performance compressor optimised for numerical
data which uses ``blocking''~\citep{alted2010modern} to 
optimise CPU-cache access patterns, as well as highly optimised
bit and byte shuffle filters.  Remarkably, on highly 
compressible datasets, Blosc decompression can be faster 
than \texttt{memcpy}.
Blosc is written in C, with APIs for C, Python, Julia, Rust
and others.
Blosc is a ``meta-compressor'' because it provides different 
access to several different compression codecs. The 
Zstandard compressor~\citep{collet2021rfc} is of particular 
interest here as it achieves very high compression ratios
with fast decompression speeds [see fig]. 
% This may seem a bit off topic, but want to reassure reader that 
% this isn't some niche technology
Zstandard is also used in several recent VCF compression 
methods~\citep[e.g.][]{lefaive2021sparse,wertenbroek2022xsi}.
% LZ4 is also notable - any citations for this?

@@

Zarr has recently been gaining popularity in biological applications.
The Open Microscopy Environment has developed OME-Zarr~\cite{moore2023ome} as one 
of its ``next generation'' cloud ready file formats~\citep{moore2021ome}.
OME-Zarr already has a rich suite of supporting
tools~\cite{moore2023ome,rzepka2023toward}. 
Zarr has also seen recent uptake 
in single-scell single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
and multimodal spatial omics
data~\citep{marconato2024spatialdata,baker2023emobject}.

EUGENe~\citep{klie2023predictive} is a toolkit for 
for applying deep-learning methods to genomic sequence data,
integrating with PyTorch [citation?].

DivBrowse~\citep{konig2023divbrowse}. Works with VCF data using 
the scikit-allel \texttt{vcf\_to\_zarr} function, (produces 
a precursor to the VCF-Zarr specification here). It's a 
browser, using Zarr as the storage backend.

Zarr was originally developed as a means efficiently accessing VCF data,
 and has been in use within the MalariaGEN project for XX years. [
ALISTAIR - details and citations here!]

@@

The specification defines a clear mapping of VCF field names (keys) to
array names, VCF Number to array shape, and VCF Type to array data type.
To take one example, consider the VCF AD genotype field defined by the
following VCF header: \texttt{\#\#FORMAT=<ID=AD,Number=A,Type=Integer,Description="Allele Depths">}

The FORMAT key \texttt{ID} maps to an array name of \texttt{call\_AD}
(FORMAT fields have a \texttt{call\_} prefix, while INFO fields have a
\texttt{variant\_} prefix; both are followed by the key name). Arrays
corresponding to FORMAT fields are 3-dimensional with shapes that look
like \texttt{(variants, samples, <Number>)} in general. In this case, the
Number A entry indicates that the field has one value per alternate allele,
which in VCF Zarr is represented as the \texttt{alt\_alleles} dimension name,
so the shape of this array is \texttt{(variants, samples, alt\_alleles)}.
The VCF Integer type can be represented as any Zarr integer type, and in
fact the specification doesn't mandate a particular integer width, so it
could be encoded as a 16 bit integer for example.
% Refer to GeL case study section?

>>>>

Zarr is flexible in allowing different compression codecs and 
pre-compression filters to be specified on a per-array basis
(see section XXX).
Two key technologies used by Zarr are the Blosc
% This seems like the best citation for Blosc? It is mentioned here
meta-compressor~\cite{alted2010modern}
and Zstandard compression algorithm~\citep{collet2021rfc}.
Blosc is a high-performance compressor optimised for numerical
data which uses ``blocking''~\citep{alted2010modern} to 
optimise CPU-cache access patterns, as well as highly optimised
bit and byte shuffle filters.  Remarkably, on highly 
compressible datasets, Blosc decompression can be faster 
than \texttt{memcpy}.
% Mentioning this detail to reassure readers who might be thinking
% "how would this be implemented and what kind of horrible dependencies
% would I need?"
Blosc is written in C, with APIs for C, Python, Julia, Rust
and others.
Blosc is a ``meta-compressor'' because it provides 
access to several different compression codecs. The 
Zstandard compressor~\citep{collet2021rfc} is of particular 
interest here as it achieves very high compression ratios
with fast decompression speeds [see fig]. 
% This may seem a bit off topic, but want to reassure reader that 
% this isn't some niche technology
Zstandard is also used in several recent VCF compression 
methods~\citep[e.g.][]{lefaive2021sparse,wertenbroek2022xsi}.
% LZ4 is also notable - any citations for this?

@@

Zarr has recently been gaining popularity in biological applications.
The Open Microscopy Environment has developed OME-Zarr~\cite{moore2023ome} as one 
of its ``next generation'' cloud ready file formats~\citep{moore2021ome}.
OME-Zarr already has a rich suite of supporting
tools~\cite{moore2023ome,rzepka2023toward}. 
Zarr has also seen recent uptake 
in single-scell single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
and multimodal spatial omics
data~\citep{marconato2024spatialdata,baker2023emobject}.
Recent additions using Zarr include the application of deep learning
models to genomic sequence data~\citep{klie2023predictive} 
and a browser for genetic variation data~\citep{konig2023divbrowse}.
Zarr has in fact been in use in genomics for many years, and was originally 
developed to [ALISTAIR - details and citations here!]

@@

The specification defines a clear mapping of VCF field names (keys) to
array names, VCF Number to array shape, and VCF Type to array data type.
To take one example, consider the VCF AD genotype field defined by the
following VCF header: \texttt{\#\#FORMAT=<ID=AD,Number=A,Type=Integer,Description="Allele Depths">}
The FORMAT key \texttt{ID} maps to an array name of \texttt{call\_AD}
(FORMAT fields have a \texttt{call\_} prefix, while INFO fields have a
\texttt{variant\_} prefix; both are followed by the key name). Arrays
corresponding to FORMAT fields are 3-dimensional with shapes that look
like \texttt{(variants, samples, <Number>)} in general. In this case, the
Number A entry indicates that the field has one value per alternate allele,
which in VCF Zarr is represented as the \texttt{alt\_alleles} dimension name,
so the shape of this array is \texttt{(variants, samples, alt\_alleles)}.
The VCF Integer type can be represented as any Zarr integer type, and in
fact the specification doesn't mandate a particular integer width, so it
could be encoded as a 16 bit integer for example.
% Refer to GeL case study section?
